### 2020-02-20 chapters/background.tex
Bloom Filter

An aggregation --- also known as a fold, reduce or accumulate --- is an operation that reduces a collection of values to a single one. Examples of such are the computation of data set sums, minima, averages or the number of distinct elements (cardinality). An aggregation may be exact or approximate. Exact aggregations have an exact value or state in that there is no error or loss of resolution associated. Approximate aggregations have an error margin to the associated aggregation value. This error is usually controllable by parameter optimization.

Aggregate operations are present in most programming languages and frameworks. Quoting from the JavaSE13 documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}: \textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}. Hadoop's MapReduce paradigm consists of first applying a map function to the collection of data and then a reduce one. The map task will transform each item as desired and the reduce task will merge the collection into a single value. Reduce methods make use of a \textit{combine} function and recursively apply it. The \textit{combine} function takes the current aggregation value and combines it with the next item in the set, returning the new aggregation state. Doing this recursively for each item in the collection results in a single value: the aggregation value for the data set.

In stream processing, aggregation plays a central role, but unlike many other scenarios that focus on data at rest, streaming aggregations handle data in motion. While some aggregations can be computed on an infinite data stream, often we want to compute these over a finite data set. As we have seen in Section \ref{sec:windows}, to obtain a bounded data set from a data stream we apply windowing techniques, thus being able to aggregate the data in the sliding window. The authors of \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs). Once again let us consider the data center temperature monitoring problem from Section \ref{sec:windows} as an example use-case for SWAGs. How would we compute the average temperature per rack in the data center in the last 5 minutes? To better understand and later on examine this aggregation, let us translate this question to a streaming SQL query, with a similar syntax to the API provided by Apache Flink \cite{flink}. Assuming all the temperatures can be queried from the \textit{Temperatures} table and that to each rack corresponds a different \textit{rack\_id} we have the following query:

In the query above, the \textit{TUMBLE(timestamp, Time.minutes(5), rack\_id)} statement groups incoming data using time-based tumbling windows of 5 minutes duration, using the \textit{timestamp} field to filter out events not in this time interval. The resulting list of rows has \textit{rack\_id} as its key and as corresponding value a list of temperatures, from the past 5 minutes. Finally, for each rack and its list of temperatures, we compute an \textit{AVG} aggregation, resulting in the average temperature per rack in the past 5 minutes.

@@

Different aggregations have different properties. These properties are listed in \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}, where the authors group different aggregations according to the properties they satisfy. Table \ref{tbl:aggregations-properties} was adapted from this work and categorizes different families of operations based on a set of properties.

\begin{table}[!htb]
    \begin{center}
      \includegraphics[scale=0.5]{figures/aggregation-operations-properties.png}
      \caption[Aggregation operations and properties]{Aggregation operations and properties. Checkmarks (\checkmark), crosses ($\times$), and question marks (?) indicate a property is true for all, false for all, or false for some of a given group of like operations, respectively}
      \label{tbl:aggregations-properties}
    \end{center}
\end{table}

In order to properly explain these properties it is necessary to first define the \textit{combine} and \textit{lower} functions. The \textit{combine} function merges the incoming event with the incrementally computed aggregation state. The \textit{lower} function is responsible for returning a result for a given query according to the current aggregation state. For instance, consider the use case of computing a sum aggregation over a window of data. The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.} the current sum value --- with the new incoming event. The \textit{lower} function would return the current aggregation state --- \textit{i.e.} return the total sum computed.

@@

\subsection{Sliding Window Aggregation Algorithms}

% Generic SWAG algorithms - Generic algorithms for the computation of exact or approximate aggregations over a data stream in a sliding window fashion (DABA, Subtract On Evict, Recalculate From Scratch, Two Stacks)
% Aggregations - reduce a collection of values to one (sum, min, avg, cardinality of a set, membership query)
  % Exact Aggregations - an aggregation whose value is exact, in that there is no error or loss of resolution associated with the computation
  % Approximate Aggregations -  an aggregation whose value has some error margin. Usually controllable.

A sliding window aggregation (SWAG) algorithm allows the computation of an exact or approximate aggregation over a sliding window. A SWAG algorithm may restrict the set of possible aggregations based on their properties. Different SWAG algorithms use different data structures for the window contents and the incremental aggregation state. Despite storing information using different structures, every SWAG algorithm must implement three methods:

@@

\subsubsection{Subtract On Evict}
Subtract On Evict (SOE) only works for aggregations whose  \textit{combine} function is invertible. That is, as defined in Table \ref{tbl:aggregations-properties}, a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$.

@@

\subsection{Summary}
This subsection adds nothing new to the previous ones. However, it concisely summarizes the most relevant previously discussed concepts.

An aggregation is an operation that reduces a collection of values to a single one. Examples of such are sums, minima, averages or finding the number of distinct elements in a set (cardinality). An aggregation may be exact or approximate. Exact aggregations have an exact value or state in that there is no error or loss of resolution associated. Approximate aggregations have an error margin to the associated aggregation value. This error is usually controllable by parameter optimization.

Sliding Window Aggregation algorithms allow the computation of exact or approximate aggregations over a data stream in a sliding window fashion. Each algorithm must implement three methods: \textit{insert}, \textit{evict} and \textit{query}. A SWAG algorithm may restrict the set of possible aggregations based on their properties.

>>>>

An aggregation --- also known as a fold, reduce or accumulate --- is an operation that reduces a collection of values to a single one. Examples of such are the computation of data set sums, minima, averages or the number of distinct elements (cardinality). An aggregation may be exact or approximate. Exact aggregations compute a value from a sequence of elements, where no error or loss of resolution exists. On the other hand, approximate aggregations have an error margin to the associated aggregation value which is usually controllable by adjusting a set of parameters.

Aggregation is present in most programming languages and frameworks. Quoting from the JavaSE13 documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}: \textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}. Hadoop's MapReduce paradigm consists of first applying \textit{map} steps to the collection of data and then a \textit{reduce} one. \textit{Map} task will transform each item as desired and the \textit{reduce} task will merge the collection into a single value. Reduce methods make use of a \textit{combine} function and recursively apply it. The \textit{combine} function takes the current aggregation value and combines it with the next item in the set, returning the new aggregated state. Doing this recursively for each item in the collection results in a single value: the aggregate value for the data set.

@@

Different aggregations have different properties. These properties are listed in \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}, where the authors group different aggregations according to the properties they satisfy. Table \ref{tbl:aggregations-properties} was adapted from this work and categorizes different families of operations based on said properties.

\begin{table}[!htb]
    \begin{center}
      \includegraphics[scale=0.5]{figures/aggregation-operations-properties.png}
      \caption[Aggregation properties]{Aggregation properties. Checkmarks (\checkmark), crosses ($\times$), and question marks (?) indicate a property is true for all, false for all, or false for some of a given group of like operations, respectively}
      \label{tbl:aggregations-properties}
    \end{center}
\end{table}

To understand these properties, we must be familiar with the \textit{combine} and \textit{lower} functions. The already mentioned \textit{combine} function merges the incoming event with the incrementally computed aggregation state. The \textit{lower} function is responsible for returning a result for a given query according to the current aggregation state. For instance, consider the use case of computing a sum aggregation over a window of data. The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.} the current sum value --- with the new incoming event. The \textit{lower} function would return the current aggregation state --- \textit{i.e.} return the total sum computed.

@@

Unlike many other scenarios that focus on data at rest, streaming aggregations handle data in motion. While some aggregations can be computed on an infinite data set, often we want to compute these over a finite one. As we have seen in Section \ref{sec:windows}, to obtain a bounded data set from a data stream we apply windowing techniques. In doing so, we obtain a finite data set denoted as a sliding window and compute aggregations over it. The authors of \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs). Once again let us consider the data center temperature monitoring problem from Section \ref{sec:windows} as an example use-case for SWAGs. How would we compute the average temperature per rack in the data center in the last 5 minutes? To better understand, let us translate this question to a streaming SQL query, with a similar syntax to the API provided by Apache Flink \cite{flink}. Assuming all the temperatures can be queried from the \textit{Temperatures} table and that to each rack corresponds a different \textit{rack\_id} we have the following query:

In the query above, the \textit{TUMBLE(timestamp, Time.minutes(5), rack\_id)} statement groups incoming data using time-based tumbling windows of 5 minutes duration, using the \textit{timestamp} field to evict events not in this time interval. The resulting list of rows has \textit{rack\_id} as its key and as corresponding value a list of temperatures, from the past 5 minutes. Finally, for each rack and its list of temperatures, we compute an \textit{AVG} aggregation, resulting in the average temperature per rack in the past 5 minutes.

\subsection{Sliding Window Aggregation Algorithms} \label{sec:back-swag-algs}

A sliding window aggregation (SWAG) algorithm allows the computation of an exact or approximate aggregation over a sliding window. A SWAG algorithm may restrict the set of possible to compute aggregations based on their properties. Different SWAG algorithms use different data structures for the window contents and the incremental aggregation state. Despite storing information using different structures, every SWAG algorithm must implement three methods:

@@

\subsubsection{Subtract On Evict}
Subtract On Evict (SOE) only works for aggregations whose  \textit{combine} $\oplus$ function is invertible. That is, as defined in Table \ref{tbl:aggregations-properties}, a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$.

@@

### 2020-02-20 chapters/sota.tex
Bloom Filter

\subsubsection{Time complexity analysis}
%k to initialize and when change is detected
%k * d
%d 
 %nlog(2k) to initialize tree and when change is detected
 %log(2k) to maintain the tree
The algorithm begins by collecting the first \textit{k} tuples. Both windows will have size \textit{k}. This takes \textit{O(k)} time but is a task that will only occur on initialization and on change detection, upon which the window contents must reset, as described in \ref{fig:change-detection-2-windows}. For each of the next \textit{k} iterations the function \textit{d} will be computed and compared to the threshold $\alpha$. As previously mentioned, computing the value of \textit{d} can be done in \textit{O(log(2k))} time per iteration. However, the initial time cost of initializing the balanced tree is of \textit{O(k log(2k))}.

@@

@@

@@

\subsubsection{Time complexity analysis}
Extra-N only performs range queries for incoming data points. This means that for each new data point \textit{p\textsubscript{i}}, Extra-N will search the area of diameter $\theta\textsuperscript{range}$ around it and update on it neighbors. In conclusion, the time complexity of Extra-N is \textit{O(n)} where \textit{n} is the number of new data points.

\subsubsection{Space complexity analysis}
In summary, Extra-N keeps the memory consumption linear to the number of data points in the window without performing unnecessary range queries.

\subsubsection{Applicability to our Hypothesis}
While a linear complexity fits most of the scenarios, it does not fit ours since we aim to build a low memory footprint solution, ideally with space complexity of \textit{O(1)}. Hence, this algorithm will be of no use to us.

@@

\subsubsection{Applicability to our Hypothesis}
Online learning algorithms that adapt to data pattern shifts and eventually consider them regular patterns are not viable methods to accomplish our goal of building a monitoring system that alerts data pattern shifts continuously until they match the initial configuration again. Additionally, insights produced by machine learning models are not human-understandable, rendering the user unable to understand why an anomaly report was made. In conclusion, we do not deem HTM applicable to our solution.

@@

\iffalse

\section{Sliding Window Aggregations}
Aggregations and Sliding Window Aggregations (SWAGs) have been introduced in Section \ref{sec:aggregations}. In this Section, we analyze algorithms that allow for sliding window aggregations utilizing a set of different aggregation operators. In other words, different algorithms work with different aggregation categories based on their properties. The properties of different categories of aggregations were defined in Section \ref{sec:aggregations} and summarized in Table \ref{tbl:operations-properties}.

\subsection{DABA}
\subsubsection{Time complexity analysis}
\subsubsection{Space complexity analysis}
\subsubsection{Applicability to our Hypothesis}

\section{Probabilistic Data Structures} \label{sec:pds}
Probabilistic data structures use hash functions to compactly represent a set of items while approximately answering queries with fixed error bounds. This way, these structures are capable of processing huge volumes of data. They require a single pass through the data, which is appropriate for a streaming scenario. Probabilistic data structures have constant space and time complexity \cite{Singh-PDS-BIGD} making them candidate methods to build our real-time and low memory footprint system.

\subsection{Membership Queries and the Bloom Filter}
A Bloom filter \cite{BLOOM-BLOOMFILTER} is a probabilistic data structure that allows membership queries on a set of elements. Bloom filters answer the query: "Is element \textit{el} in the set of seen elements so far?". Being a probabilistic data structure, Bloom filters do not always give a certain answer. \textit{False positive} matches are possible --- \textit{i.e.} saying \textit{el} is in the set when it is not --- but \textit{false negatives} are not --- \textit{i.e.} saying \textit{el} is not in the set but it actually is. In other words, a bloom filter will accurately identify all items that do not belong in the set but will misclassify some items as being present in the set. Hence, a query returns either \textit{possibly in set} or \textit{definitely not in set}.

A Bloom filter is an array of bits of size \textit{m}. A Bloom Filter makes use of \textit{k} hash functions. The choice of the \textit{m} and \textit{k} constants will determine the false positive rate.

Initially, all bits are set to 0. Adding an element is done by determining which positions should be set to 1. To that end, the new element is fed into each hash function that maps the element to an array position. The resultant \textit{k} outputs are used as the \textit{k} positions of the array to be set to 1.

Performing a membership query --- \textit{i.e.} testing if an element \textit{el} is in the set --- is done by feeding \textit{el} into each one of the \textit{k} hash functions in order to get an array of positions. If \textit{el} was already in the set then all \textit{k} positions should be 1. If any position contains a 0 then the element is \textit{definitely not in set}, as shown in Figure \ref{fig:bloom-filter} from \cite{bloom-filter-wikipedia}. If all positions are 1 then the element is \textit{possibly in the set}.

The reason a Bloom Filter does not provide 100\% certainty that \textit{el} is in the set in case of all positions being 1 is that hash functions may give the same position for two different elements. For example, hash functions \textit{k\textsubscript{1}}, \textit{k\textsubscript{2}} and \textit{k\textsubscript{3}} may map an element \textit{el1} to positions $[1,2,3]$ and an element \textit{el2} to positions $[4,5,6]$. This way, bits in positions $[1,2,3,4,5,6]$ are set to 1. Given a new element \textit{el3} not yet seen, hash functions \textit{k\textsubscript{1}}, \textit{k\textsubscript{2}} and \textit{k\textsubscript{3}} may map it to $[1,3,5]$ and all of these positions are already 1 because of the previous elements. Thus the Bloom filter returns \textit{"possibly in set"} because all bits are set to 1 when in reality \textit{el3} was not in the set. This is considered a \textit{false positive}.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.8]{figures/bloom-filter.png}
      \caption[Bloom filter membership test]{Bloom filter membership test of element 'W'. Finding one '0' indicates that it is not in the set.}
      \label{fig:bloom-filter}
    \end{center}
\end{figure}

An improvement to this structure was defined in \cite{Kirsch-Better-Bloom} where the authors attempt to reduce the overhead of bloom filters operations such as adding a new element or answering a query by focusing on the hash functions. The proposed method is based on hashing literature. It has been shown that multiple hash functions can be created from the combination of only two hash functions. This means that a new hash function \textit{g} can be produced based on  two other independent and uniform existing hash functions, \textit{h1} and \textit{h2}. The new hash function will map elements to a universe ranging from \textit{0} to \textit{p-1} and will be defined as \textit{$g(x) = h1(x) + ih2(x) mod p$}. Since the only nontrivial computations performed by a Bloom filter are the evaluations of pseudo-random hash functions, any reduction in the required number of pseudo-random hash functions yields a reduction in the time required to add an element or query a Bloom filter.

Bloom filters are not the only approximate aggregators that use hash functions as a basis of their algorithm. All of the studied probabilistic data structures do so. Therefore, the technique proposed is beneficial in HyperLogLogs, Count-Min Sketches and Bloom filters.

\subsubsection{Time complexity analysis}
Considering a Bloom filter with \textit{m} bits and \textit{k} hash functions, insertion and search will both take \textit{O(k)} time because all there is to do is run the input through all of the \textit{k} hash functions and set the bits in the given positions to 1. Note the time complexity does not at all depend on the number of elements in it and that \textit{k} will be a rather small constant (magnitude of $10^1$). Hence, the Bloom filter is said to have constant time complexity (\textit{O(1)}).

\subsubsection{Space complexity analysis}
Considering a Bloom filter with \textit{m} bits, the space required is simply the array of \textit{m} bits, thus \textit{O(m)} space complexity. Similarly to the time complexity analysis, we point out that \textit{m} will be constant and that each of the array elements occupies 1 bit. Thus the Bloom filter has a constant space complexity (\textit{O(1)}).

\subsubsection{Applicability to our Hypothesis}
Bloom filters exhibit constant space and time complexity making it great methods to test our hypothesis. Their false positive rate is tolerable in most scenarios and can be controlled by tweaking the values of \textit{m} and \textit{k}.

@@

\subsubsection{Time complexity analysis}
When adding an element or querying the frequency of one, the Count-Min Sketch loops through each row and applies a constant in time hash function, proceeding to a position update or retrieval, respectively. Hence, the time complexity is of \textit{O(d)}. Given \textit{d} is a constant, the CMS has constant time complexity --- \textit{O(1)}.

\subsubsection{Space complexity analysis}
The Count-Min Sketch relies on a two-dimensional matrix as its only data structure. The memory used will correspond to the \textit{wd} counts, hence \textit{O(wd)} space complexity. Since both \textit{w} and \textit{d} are constants, the CMS has constant space complexity --- \textit{O(1)}.

\subsubsection{Applicability to our Hypothesis}
Count-Min Sketches are constant in both time and space. Similarly to Bloom filters, this makes them ideal candidates for our solution.

@@

\subsubsection{Time complexity analysis}
Adding an item is tantamount to hashing it, determining the corresponding bucket based on the first \textit{n} bits, count the leftmost 1-bit position and update the bucket value if need be. Estimating the cardinality of the set amounts to finding the maximum value \textit{p} between all buckets and computing $2^p$. Hence, both inserting a new element and estimating the cardinality take constant time --- \textit{O(1)}.

\subsubsection{Space complexity analysis}
Quoting Flajolet et al. about the memory consumption and error in the approximations obtained in their experiments in  \cite{Flajolet-HLL}: 
\say{cardinalities till values over N = $10^9$ can be estimated with a typical accuracy of 2\% using 1.5kB (kilobyte) of storage.}

@@

\subsubsection{Applicability to our Hypothesis}
Similar to the previously studied probabilistic data structures, HyperLogLog (HLL) has constant space and time complexity while giving an approximation of a set's cardinality. Hence, HLLs are considered valuable methods for our work.

\iffalse

\section{Sliding Window Aggregation with Probabilistic Data Structures}
In Section \ref{sec:pds} we presented a set of probabilistic data structures constant in time and space complexity. These structures aggregate an endless stream of data. However, when working under a sliding window context there is the need not only to insert new events but to evict old ones. Probabilistic data structures aggregate all incoming data but have no way of removing elements.

>>>>

\subsubsection*{Time complexity analysis}
%k to initialize and when change is detected
%k * d
%d 
 %nlog(2k) to initialize tree and when change is detected
 %log(2k) to maintain the tree
The algorithm begins by collecting the first \textit{k} tuples. Both windows will have size \textit{k}. This takes \textit{O(k)} time but is a task that will only occur on initialization and on change detection, upon which the window contents must reset, as described in \ref{fig:change-detection-2-windows}. For each of the next \textit{k} iterations the function \textit{d} will be computed and compared to the threshold $\alpha$. As previously mentioned, computing the value of \textit{d} can be done in \textit{O(log(2k))} time per iteration. However, the initial time cost of initializing the balanced tree is of \textit{O(k log(2k))}.

@@

\subsubsection*{Space complexity analysis}

@@

\subsubsection*{Applicability to our Hypothesis}

@@

\subsubsection*{Time complexity analysis}
Extra-N only performs range queries for incoming data points. This means that for each new data point \textit{p\textsubscript{i}}, Extra-N will search the area of diameter $\theta\textsuperscript{range}$ around it and update on it neighbors. In conclusion, the time complexity of Extra-N is \textit{O(n)} where \textit{n} is the number of new data points.

\subsubsection*{Space complexity analysis}
In summary, Extra-N keeps the memory consumption linear to the number of data points in the window without performing unnecessary range queries.

\subsubsection*{Applicability to our Hypothesis}
While a linear complexity fits most of the scenarios, it does not fit ours since we aim to build a low memory footprint solution, ideally with space complexity of \textit{O(1)}. Hence, this algorithm will be of no use to us.

@@

\subsubsection*{Applicability to our Hypothesis}
Online learning algorithms that adapt to data pattern shifts and eventually consider them regular patterns are not viable methods to accomplish our goal of building a monitoring system that alerts data pattern shifts continuously until they match the initial configuration again. Additionally, insights produced by machine learning models are not human-understandable, rendering the user unable to understand why an anomaly report was made. In conclusion, we do not deem HTM applicable to our solution.

@@

\section{Sliding Window Aggregation Algorithms} \label{sec:sota-swag-algs}
As discussed in Section \ref{sec:back-swag-algs} ...
Besides RFS and SOE other methods were proposed like DABA...
\subsection{DABA}
\subsubsection*{Time complexity analysis}
\subsubsection*{Space complexity analysis}
\subsubsection*{Applicability to our Hypothesis}

\section{Probabilistic Data Structures} \label{sec:pds}
In Section \ref{sec:sota-swag-algs} we presented some state-of-the-art algorithms for computing sliding window aggregations (SWAGs). However, if the aggregation we desire to compute can be approximated, there may exist more memory efficient methods. These aggregators are commonly denoted as sketches or probabilistic data structures and produce results orders-of magnitude faster then some exact approaches while keeping the memory consumption constant and providing mathematically proven error bounds. Each probabilistic data structure is used for efficient computation of one approximate aggregation, hence we say they are aggregation-specific.

Probabilistic data structures use hash functions to compactly represent a set of items. They require a single pass through the data, which is appropriate for a streaming scenario. Probabilistic data structures have constant space and time complexity \cite{Singh-PDS-BIGD} making them ideal aggregators to build our real-time and low memory footprint system.

% FOR EACH SKETCH DO:
% approx aggregation to compute
% data structures
% algorithm for insertion
% algorithm for query
% error bound related with parameters
% time complexity analysis
% space complexity analysis
% Applicability to our Hypothesis

\subsection{Membership Queries and the Bloom Filter}
% approx aggregation to compute
A Bloom filter \cite{BLOOM-BLOOMFILTER} aggregator allows the computation of a membership query approximate aggregation on a set of elements. Bloom filters answer the query: "Is element \textit{el} in the set of seen elements so far?". Being a probabilistic data structure and approximate aggregator, Bloom filters' answer has an error associated. \textit{False positive} matches are possible --- \textit{i.e.} saying \textit{el} is in the set when it is not --- but \textit{false negatives} are not --- \textit{i.e.} saying \textit{el} is not in the set but it actually is. In other words, a bloom filter will accurately identify all items that do not belong in the set but will misclassify some items as being present in the set. Hence, a query returns either \textit{possibly in set} or \textit{definitely not in set}.

% data structures
\subsubsection*{Data structure}
A Bloom filter is an array of bits of size \textit{m}, initially all set to 0. Figure \ref{fig:initial-bloom-filter} represents a Bloom filter with a bit array of size \textit{m}=7 with all bits initially set to 0.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=1.1]{figures/initial-bloom.png}
      \caption[Bloom filter bit array initial state]{Bloom filter bit array initial state of \textit{m}=7}
      \label{fig:initial-bloom-filter}
    \end{center}
\end{figure}

% algorithm for insertion
\subsubsection*{Element Insertion}
Element insertion and query answering make use of \textit{k} hash functions, each mapping elements to a position in the bit array. Adding an element is done by determining which positions should be set to 1. To that end, the new element is fed into each hash function that maps the element to an array position. The resultant \textit{k} outputs are used as the \textit{k} positions of the array to be set to 1.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/insert-bloom.png}
      \caption[Bloom filter insertion]{Insertion of element \textit{el} into a Bloom filter with \textit{m}=7 and \textit{k}=3}
      \label{fig:insertion-bloom-filter}
    \end{center}
\end{figure}

% algorithm for query
\subsubsection*{Membership Query}
Performing a membership query --- \textit{i.e.} testing if an element \textit{el} is in the set --- is done by feeding \textit{el} into each one of the \textit{k} hash functions in order to get an array of positions. If \textit{el} was already in the set, all \textit{k} positions should be 1. If any position contains a 0 then the element is \textit{definitely not in set}, as shown in Figure \ref{fig:bloom-filter}.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/query-bloom.png}
      \caption[Bloom filter membership test]{Bloom filter membership test of element \textit{el}. Finding at least one '0' indicates that it is not in the set.}
      \label{fig:bloom-filter}
    \end{center}
\end{figure}

However, if all positions are 1 then the element is said to be \textit{possibly in the set}. The reason a Bloom Filter does not provide 100\% certainty that \textit{el} is in the set in case of all positions being 1 is that hash functions may give the same position for two different elements. For example, hash functions \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}} may map an element \textit{el1} to positions $[1,2,3]$ as illustrated in Figure \ref{fig:bloom-filter-fp1} and an element \textit{el2} to positions $[4,5,6]$ like in Figure \ref{fig:bloom-filter-fp2}. This way, bits in positions $[1,2,3,4,5,6]$ are set to 1. When querying about the membership of a new element \textit{el3}, hash functions \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}} may map it to positions $[1,3,5]$ as exemplified in Figure \ref{fig:bloom-filter-fp3}. Since all of these positions are 1 because of the previous elements inserted, the Bloom filter returns \textit{"possibly in set"} when in reality \textit{el3} was not in the set. This is considered a \textit{false positive}.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/fp-bloom-1.png}
      \caption[Bloom filter insertion]{Insertion of \textit{el1} in the Bloom filter}
      \label{fig:bloom-filter-fp1}
    \end{center}
\end{figure}

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/fp-bloom-2.png}
      \caption[Bloom filter insertion]{Insertion of \textit{el2} in the Bloom filter}
      \label{fig:bloom-filter-fp2}
    \end{center}
\end{figure}

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/fp-bloom-3.png}
      \caption[Bloom filter false positive]{Membership query of \textit{el3} and false positive result}
      \label{fig:bloom-filter-fp3}
    \end{center}
\end{figure}

\subsubsection*{False Positive rate}
% error bound related with parameters
The false positive rate is a function of the bit array size \textit{m}, the number of hash functions \textit{k} and the number of elements in the set \textit{n}. It is assumed that the hash functions produce an independent and random index value for each element. Mullin presents an analysis of the Bloom filter in \cite{Mullin-Bloom-Analysis} where he details the false positive rate formula.

The probability of one arbitrary bit being set to 1 is
\begin{equation}
    \frac{1}{m}
\end{equation}
because there are \textit{m} bits. Likewise, the probability of one bit not being set to 1 is
\begin{equation}
    1-\frac{1}{m}
\end{equation}

Mullin states the probability that an arbitrary bit is not set to 1 after one element insertion and corresponding \textit{k} bit updates is
\begin{equation}
    (1-\frac{1}{m})^\textit{k}
\end{equation}

For \textit{n} insertions, the probability a bit is not set to 1 is
\begin{equation}
    (1-\frac{1}{m})^\textit{kn}
\end{equation}

Therefore, the probability that an arbitrary bit is set to 1 after \textit{n} insertions is 
\begin{equation}
    1-(1-\frac{1}{m})^\textit{kn}
\end{equation}

Hence, the probability of false positive of a membership query for a new element for \textit{k} hash functions is given by
\begin{equation}
    (1-(1-\frac{1}{m})^\textit{kn})^\textit{k}
\end{equation}

This false positive rate can be approximated by 
\begin{equation}
    (1-e^\textit{-kn/m})^\textit{k}
\end{equation}

\subsubsection*{Optimal number of hash functions}
We want to choose the number of hash functions \textit{k} in order to minimize the false positive rate. In order to do so, we must first allocate \textit{m} bits for the Bloom filter, choosing \textit{m} based on available memory. The value of \textit{k} that minimizes the false positive rate is either of the nearest integers given by
\begin{equation}
    \frac{m}{n}ln(2)
\end{equation}

\iffalse
%TODO hash improvement probably better declared outside of bloom
An improvement to this structure was defined in \cite{Kirsch-Better-Bloom} where the authors attempt to reduce the overhead of bloom filters operations such as adding a new element or answering a query by focusing on the hash functions. The proposed method is based on hashing literature. It has been shown that multiple hash functions can be created from the combination of only two hash functions. This means that a new hash function \textit{g} can be produced based on  two other independent and uniform existing hash functions, \textit{h1} and \textit{h2}. The new hash function will map elements to a universe ranging from \textit{0} to \textit{p-1} and will be defined as \textit{$g(x) = h1(x) + ih2(x) mod p$}. Since the only nontrivial computations performed by a Bloom filter are the evaluations of pseudo-random hash functions, any reduction in the required number of pseudo-random hash functions yields a reduction in the time required to add an element or query a Bloom filter.

Bloom filters are not the only approximate aggregators that use hash functions as a basis of their algorithm. All of the studied probabilistic data structures do so. Therefore, the technique proposed is beneficial in HyperLogLogs, Count-Min Sketches and Bloom filters.
\fi

\subsubsection*{Time complexity analysis}
Considering a Bloom filter with \textit{m} bits and \textit{k} hash functions, insertion has \textit{O(k)} time complexity because all there is to do is run the input through all of the \textit{k} hash functions and set the bits in the given positions to 1. Similarly, query answering has \textit{O(k)} time complexity because it needs only to run the input through the \textit{k} hash functions and check if all bits are set to 1. If so, it returns \textit{possibly in set}. Otherwise it returns \textit{definitely not in set}.

Note the time complexity does not at all depend on the number of elements in it and that \textit{k} will be a rather small constant. Hence, the Bloom filter is said to have constant time complexity.

\subsubsection*{Space complexity analysis}
Considering a Bloom filter with \textit{m} bits, the space required is simply the array of \textit{m} bits, thus \textit{O(m)} space complexity. Similarly to the time complexity analysis, we point out that \textit{m} will be constant and that each of the array elements occupies 1 bit. Thus the Bloom filter has a constant space complexity.

\subsubsection*{Applicability to our Hypothesis}
Bloom filters exhibit constant space and time complexity making it great methods to test our hypothesis. Their false positive rate is tolerable in most scenarios and can be controlled by tweaking the values of \textit{m} and \textit{k}.

@@

\subsubsection*{Time complexity analysis}
When adding an element or querying the frequency of one, the Count-Min Sketch loops through each row and applies a constant in time hash function, proceeding to a position update or retrieval, respectively. Hence, the time complexity is of \textit{O(d)}. Given \textit{d} is a constant, the CMS has constant time complexity --- \textit{O(1)}.

\subsubsection*{Space complexity analysis}
The Count-Min Sketch relies on a two-dimensional matrix as its only data structure. The memory used will correspond to the \textit{wd} counts, hence \textit{O(wd)} space complexity. Since both \textit{w} and \textit{d} are constants, the CMS has constant space complexity --- \textit{O(1)}.

\subsubsection*{Applicability to our Hypothesis}
Count-Min Sketches are constant in both time and space. Similarly to Bloom filters, this makes them ideal candidates for our solution.

@@

\subsubsection*{Time complexity analysis}
Adding an item is tantamount to hashing it, determining the corresponding bucket based on the first \textit{n} bits, count the leftmost 1-bit position and update the bucket value if need be. Estimating the cardinality of the set amounts to finding the maximum value \textit{p} between all buckets and computing $2^p$. Hence, both inserting a new element and estimating the cardinality take constant time --- \textit{O(1)}.

\subsubsection*{Space complexity analysis}
Quoting Flajolet et al. about the memory consumption and error in the approximations obtained in their experiments in  \cite{Flajolet-HLL}: 
\say{cardinalities till values over N = $10^9$ can be estimated with a typical accuracy of 2\% using 1.5kB (kilobyte) of storage.}

@@

\subsubsection*{Applicability to our Hypothesis}
Similar to the previously studied probabilistic data structures, HyperLogLog (HLL) has constant space and time complexity while giving an approximation of a set's cardinality. Hence, HLLs are considered valuable methods for our work.

% PDS have no way to expire old data (must for SWAG)
% they could be implemented using one of the previously seen SWAG algorithms but lose the space efficiency
% present sliding versions as solution

\iffalse
\section{Sliding Window Aggregation with Probabilistic Data Structures}
In Section \ref{sec:pds} we presented a set of probabilistic data structures constant in time and space complexity. These structures aggregate an endless stream of data. However, when working under a sliding window context there is the need not only to insert new events but to evict old ones. Probabilistic data structures aggregate all incoming data but have no way of removing elements.
